---
title: "TnSeq -- mapping barcoded transposons to genome"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook: 
    theme: spacelab
    toc: yes
---

## Description

This R notebook is a bioinformatics pipeline to map reads from a barcoded transposon library to the genome of a target organism. For background and details regarding the method, see [Wetmore at al., mBio, 2015](https://mbio.asm.org/content/6/3/e00306-15) and [Price et al., Nature, 2018](http://www.nature.com/articles/s41586-018-0124-0)). The initial steps of processing next generation sequencing data was directly adapted from [Morgan Price's Feba repository](https://bitbucket.org/berkeleylab/feba/src/master/), see also `README.md` of this repository.

## Libraries

```{r, message = FALSE}
library(lattice)
library(latticeExtra)
library(latticetools)
library(tidyverse)
library(stringi)
```

## Annotation of mapped reads

As a prerequisite to assign genome-mapped reads to genes (or other coding regions of interest), we have to **prepare a gene table first**. This table serves as input for the script `DesignRandomPool.pl`, see `README.md` for this github repository. This table is nothing more than a slightly trimmed version of the (RefSeq) `*.gff` file that can be downloaded from NCBI genome.

All reference genome tables were already downloaded in `ref/`, and can be imported from there.

```{r, message = FALSE, warning = FALSE}
# get file names
filenames = list.files("../ref/", pattern = ".gff$")

# apply trimming function
df_ref <- lapply(filenames, FUN = function(filename) {
    
    # read tables
    read_tsv(paste0("../ref/", filename), skip = 9, col_names = FALSE) %>%
    
    # remove unnecessary rows and columns
    filter(X3 != "CDS", X3 != "region") %>%
    select(X1, X4, X5, X7, X3, X9) %>%
    
    # set column names
    rename_with(~ c("scaffoldId", "begin", "end", "strand", "desc", "locus_tag")) %>%
    
    # extract old and new locustag
    mutate(
      old_locus_tag = stri_extract_first(locus_tag, regex = "H16_[AB][0-9]*|PHG[0-9]*|HPF_[0-9]{5}"),
      new_locus_tag = stri_extract_first(locus_tag, regex = "(H16|HPF)_RS[0-9]{5}")
    ) %>%
    
    # remove old ID column
    select(-locus_tag)
  })

lapply(df_ref, head)
```

Export the modified tables to `ref/` folder.

```{r}
# assign new filenames
names(df_ref) <- gsub(".gff$", "_trimmed.tsv", filenames)

# and save to disk
silent <- lapply(names(df_ref), function(filename) {
  write_tsv(df_ref[[filename]], path = paste0("../ref/", filename))
})
```

After this, the actions outlined in `README.md` can be performed. This is mainly A) mapping reads to the genome and B) generating a read/barcode pool file (summary) from all mapped reads. These steps are performed using the original perl scripts from Morgan Price.

## Mapping statistics and distribution on genome

The next step is to inpsect basic statistics of transposon insertions and their distribution over the genome.

### Basic statistics

First read in pooled data table and display summary statistics. Remove barcodes without mapped position.

```{r, message = FALSE}
df_pool <- read_tsv("../data/pool/H16_barcode_pool.tsv") %>%
  filter(!is.na(pos))
```

```{r, echo = FALSE, message = FALSE}
paste("Number of unique barcodes:", nrow(df_pool))
paste("Number of barcodes with >= 10 reads:", df_pool %>% filter(nTot >= 10) %>% nrow)
paste("Number of barcodes with only 1 read:", df_pool %>% filter(nTot == 1) %>% nrow)
paste("Number of barcodes with > 1 position:", df_pool %>% filter(n2 > 0) %>% nrow)
paste("Number of barcodes on -/+ strand:", df_pool %>% group_by(strand) %>%
  summarize(n = length(n)) %>% pull(n) %>% paste(collapse = ", "))
```

Next we can plot the frequency of reads per barcoded transposons.

```{r, message = FALSE}
histogram(~ nTot | paste("strand:", strand), df_pool,
  par.settings = custom.colorblind(),
  between = list(x = 0.5, y = 0.5), xlim = c(-5, 55),
  xlab = "number of read per barcode", breaks = 20,
  scales = list(alternating = FALSE),
  panel = function(x, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.histogram(x, ...)
  }
)
```


### Distribution over the genome

```{r}
xyplot(nTot ~ pos | scaffold,
  df_pool %>% arrange(pos),
  par.settings = custom.colorblind(),
  between = list(x = 0.5, y = 0.5),
  layout = c(1,3), type = "l", lwd = 1.5,
  #ylab = expression("q"[S]*" [g h"^-1*" gDCW"^-1*"]"),
  #xlab = expression('µ [h'^'-1'*']'),
  scales = list(alternating = FALSE),
  panel = function(x, y, ...) {
    panel.grid(h = -1, v = -1, col = grey(0.9))
    panel.xyplot(x, y, ...)
  }
)
```

Note: for later use function foverlaps to map transposon to gene regions (or vice versa).